---
title: "ESM 244 Lab 1"
author: "Janelle Gaun"
date: "1/6/2022"
output: html_document
---

```{r setup, include=TRUE, warning= FALSE, message= FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warnings= FALSE)
library(tidyverse)
library(here)
library(sf)
library(tmap)
```

### Read in the data

```{r}
sf_trees <- read_csv(here('data', 'sf_trees', 'sf_trees.csv'))
```
###Part 1: Wrangling and ggplot

** Example 1** Find counts of observations from 'legal_status' and wrangle a bit. 

```{r}
### method 1: Group by() %>% summarize()

sf_trees %>% 
  group_by(legal_status) %>% 
  summarize(tree_count =n())

### Method 2: Different way plus a few more functions 

top_5_status <- sf_trees %>% 
  count(legal_status) %>% 
  drop_na(legal_status) %>% #drop_na drops any na enteries. we are specifying dropping na in the legal_status column only
  rename(tree_count = n) %>% 
  relocate(tree_count) %>% #brings table to front of the table
  slice_max(tree_count, n=5) %>% 
  arrange(-tree_count) #the negative sign indicates from highest to lowest

```


Make a graph of the top 5 from above: 

```{r}
ggplot(data=top_5_status, aes(fct_reorder(legal_status, tree_count), y = tree_count)) +
  geom_col(fill = 'darkgreen') +
  labs(x = 'Legal Status', y = "Tree Count")+ 
  coord_flip()+ 
  theme_minimal()
```

**Example 2:** Only Keep observation where legal status is 'Permitted Site' and caretaker is 'MIA, and store as a 'permitted_data_df'

shift-cmd-c to comment/uncomment quickly
```{r}
#sf_trees$legal_status %>% unique()
# unique(sf_trees$caretaker)

permitted_data_df <- sf_trees %>% 
  filter(legal_status %in% c('Permitted Site', 'Private') & caretaker %in% 'MTA')
  
```


**Example 3:** Only keep Blackwood Acacia trees, and then only keep solumns 'legal_status', 'date', 'latitude' and store as 'blackwood_acacia_df'


```{r}
blackwood_acacia_df <- sf_trees %>% 
  filter(str_detect(species, 'Blackwood Acacia')) %>% 
  select(legal_status, date, lat = latitude, lon = longitude)


##Make a little graph of locations
ggplot(data = blackwood_acacia_df, 
       aes(x = lon, y = lat)) +
  geom_point(color= 'darkgreen')
  
```

** Example 4:** Use 'tidyr::separate()''

```{r}
sf_trees_sep <- sf_trees %>% 
  separate(species,into = c('spp_scientific', 'species_common'), sep =' :: ')
```


**Example 5:** Use 'tidyr::unite()

```{r}
ex_5 <- sf_trees %>% 
    unite('id_status', tree_id, legal_status, sep = '_R_')
```









